---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spSGMCMC

<!-- badges: start -->
<!-- badges: end -->

spSGMCMC is designed for the Bayesian inference on massive geostatistical data using GP models which utilize stochastic gradient optimization instead of conventional MCMC. The package accompanies [this paper on arXiv](https://arxiv.org/abs/2405.04531)

## Installation

You can install the development version of spSGMCMC from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("reetamm/spSGMCMC")
```

## Example

This is an example based on the Argo data used in `GpGp`. You'll need the `GpGp` package for this example to run, since we also use it for our initial values.

```{r example, silent=T}
# install.packages("GpGp")
library(GpGp)
library(spSGMCMC)
library(ggplot2)
library(tidyverse)
library(viridis)
```


```{r, fig.asp = 0.8, fig.width = 7}
## Open and plot the data
argo2016 <- GpGp::argo2016
argo2016$lon <- ifelse(argo2016$lon>180,argo2016$lon-360,argo2016$lon)
world <- map_data("world")
argoplot <- ggplot() +
  geom_map(
    data = world, map = world,
    aes(long, lat, map_id = region),
    color = "gray", fill = "white", size = 0.01
  ) +
  geom_point(
    data = argo2016,
    aes(lon, lat, color = temp100),
    alpha = 0.7,size=0.1
  ) + coord_fixed()+theme_bw() +
  scale_color_viridis_c(option = "C")+
  xlab("Longitude")+
  ylab("Latitiude")+
  labs(colour = "Temp (C)")
argoplot
  ```